Detecting skill level and mental template in biface morphology: Archaeological and experimental insights

### 2022-04-21 manuscript/manuscript.Rmd
manuscript outline added.

Stone tools

# Materials and Methods

@@

The biface experimental replicas used in this study comprised two sub-collection. The first sub-collection includes 10 bifaces knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014]. The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 bifaces in total. Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given.

@@

## Lithic Analysis

@@

## Statistical Analyses

@@

## Principal Component analysis

Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather decent explained variance ratio to avoid overfitting. We then decided to focus on and further interpreted the implications of these first two components based on their relationships between variables (**Table** \@ref(tab:tab1)). The first principle component (PC1) indicates the overall biface thickness as it is positively correlated with all thickness measurements while negatively correlated with all other measurements. That being said, a higher PC1 value indicates a thicker biface, and vice versa. The second principle component (PC2) tracks the elongation and pointedness based on its positive relationship with maximum length and bottom width/thickness. As PC2 increases, a biface will be generally longer and more pointed since its bottom part will be bulkier.

```{r Screeplot, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scree plot showing the percentage of explained variances of the first 10 principle components."}
knitr::include_graphics(here::here("figure/Scree plot.png"))
```

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1<-read.csv(here::here("data/variable_loadings_pc12.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principle components") %>%
  kableExtra::kable_styling(position = "left", full_width = TRUE, latex_options = c("striped", "HOLD_position"), font_size = 8)%>%
kableExtra::row_spec(0, bold = T)

@@

The most salient finding of our exploratory study is that

@@

Our exploratory study produced findings with both theoretical and methodological implications.

>>>>

Handaxe is a well-studied and well-published topic, but some aspects of it still remain elusive. More specifically, the sources of variation of handaxe morphology can be very complex, but skill level and mental template are two major factors.

Skill level: short literature review

Mental template: short literature review

Here we have two interconnected research questions based on a reference sample generated from our 90-hour handaxe skill acquisition experiment: 1) Can skill level and mental template (or "aesthetic preference") be effeciently detected from biface morphometric data? 2) What is the effect of training on these two aspects?

# Materials and methods

@@

The biface experimental replicas used in this study comprised two sub-collection. The first sub-collection includes 10 bifaces knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014]. The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@bayani2021; @pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 bifaces in total. Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given.

@@

## Lithic analysis

@@

## Statistical analyses

@@

## Principal component analysis

Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather decent explained variance ratio to avoid overfitting. We then decided to focus on and further interpreted the implications of these first two components based on their relationships between variables (**Table** \@ref(tab:tab1)). The first principal component (PC1) indicates the overall biface thickness as it is positively correlated with all thickness measurements while negatively correlated with all other measurements. That being said, a higher PC1 value indicates a thicker biface, and vice versa. The second principal component (PC2) tracks the elongation and pointedness based on its positive relationship with maximum length and bottom width/thickness. As PC2 increases, a biface will be generally longer and more pointed since its bottom part will be bulkier.

```{r Screeplot, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scree plot showing the percentage of explained variances of the first 10 principal components."}
knitr::include_graphics(here::here("figure/Scree plot.png"))
```

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1<-read.csv(here::here("data/variable_loadings_pc12.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principal components") %>%
  kableExtra::kable_styling(position = "left", full_width = TRUE, latex_options = c("striped", "HOLD_position"), font_size = 8)%>%
kableExtra::row_spec(0, bold = T)

@@

PC1 (thickness) is a robust indicator of skill level as it is shared by modern expert knapper and boxgrove foragers, while PC2 (elongation and pointedness) reflects more of personal/community-level aesthetic choices. However, we do not intend to construct a false binary framework and put these two factors as disconnected and opposite concepts.

Training is important in terms of its immediate effects (pre-training vs. post-training) in both skill level and mental template, however, 90 hours of training is still not enough for novice to reach the reach the skill level as reflected in expert knappers or Late Acheulean toolmaker.

The pre-training group is similar to the Boxgrove group in PC2 because of the raw material restriction in experimental set up, which is typically flat flakes? Novice before training cannot effectively reduce the nodule to a pointed shape, making their products assembling Boxgrove in this dimension.

Experimental archaeology has a huge potential in similar topics.

@@

Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with thickness while the latter is reflected in elongation and pointedness; 2) Training has an effect, but 90 hours of training is still not enough for novice to reach the level of expertise.
